---
title: "ArgActivity"
author: "Nicole Davis"
date: "11/6/2018"
output: html_document
editor_options: 
  chunk_output_type: console
---

#prepare workspace
```{r, include=FALSE}
#load needed packages
library(readxl)
library(psych)
library(reshape2)
library(ggplot2)
library(dplyr)
library(openxlsx)
library(ggrepel)
library(knitr)
library(stringr)
library(cowplot)


redbluePalette <- c('red4','red3','red1','pink','steelblue1','royalblue2','mediumblue','navy', 'gray')

resilience_levels <- c('PWK/PhJ','A/J','NOD/ShiLtJ','CAST/EiJ','C57BL/6','129S1/SvImJ','NZO/HILtJ','WSB/EiJ', 'DO')

#set working directory
setwd('/Users/nicole/Desktop/Science/Schneider Lab/Malaria/Arginine/Arginase activity assay/')

theme_set(theme_cowplot(font_size=10))
```

#load and merge data
```{r}
aaa <- data.frame(read.xlsx('/Users/nicole/Desktop/Science/Schneider Lab/Malaria/Arginine/Arginase activity assay/20201101_ArgActivityAssay_strainsleftovers.xlsx', sheet='abs'))

aaa_meta <- data.frame(read.xlsx('/Users/nicole/Desktop/Science/Schneider Lab/Malaria/Arginine/Arginase activity assay/20201101_ArgActivityAssay_strainsleftovers.xlsx', sheet='Metadata'))

aaa_meta2 <- data.frame(read.xlsx('/Users/nicole/Desktop/Science/Schneider Lab/Malaria/_MouseStrains_x_Metabolites/MouseStrains_x_Metabolites_withliver.xlsx'))
aaa_meta2 <- unique(aaa_meta2[c('SAMPLE.NUMBER.MASTER','Mouse.ID')])

aaa_meta <- merge(aaa_meta, aaa_meta2, by.x = 'num', 'SAMPLE.NUMBER.MASTER', all.x = T)

aaa <- melt(aaa, id.vars = 'Time', variable.name = 'Well', value.name = 'Abs570')

aaa <- merge(aaa, aaa_meta, by.x = 'Well', by.y = 'Well')
aaa$Mouse.ID <- ifelse(aaa$sampletype %in% c('standard','control'), yes = aaa$num, no = aaa$Mouse.ID) #since there is only the 'num' column filled out from excel spreadsheet, need to put info for standards and controls into the Mouse.ID column
```

##wrangle data
#this chunk adds one extra piece of metadata for the standards samples
```{r}
#make metadata data - the concentration of h2o2 in the standards
h2o2 <- data.frame(Standard = paste0('S',1:6),
                   h2o2_conc = c(0, 2, 4, 6, 8, 10))

#choose column order and merge
col_order <- colnames(aaa)
aaa <- merge(aaa, h2o2, by.x = 'Mouse.ID', by.y = 'Standard', all.x = TRUE)
aaa <- aaa[c(col_order, 'h2o2_conc')]
```

##calculate things
```{r}
# #get mean of standards and samples (n=1 for samples, so just the value)
# aaa <- aaa %>%
#   group_by(Time, Mouse.ID, welltype) %>%
#   mutate(meanabs = mean(Abs570))

#plot the standards over time so you can decide how to substract the background
#if the background is constant, great; average all background values and subtract that
#otherwise, group by time and subtract background from time-matched samples
gg <- ggplot(aaa[aaa$Mouse.ID == 'S1',], aes(x=Time, y=Abs570))
gg + geom_point()
#looks like it'll be only mildly important to group by time, but to be safe, let's do it

bground <- aaa$Abs570[aaa$Mouse.ID=='S1' & aaa$Time == 0][1]

#get corrected value by subtracting background from all samples, controls, standards
aaa <- aaa %>%
  mutate(corrected = Abs570 - bground)
```

#rough plot & background-corrected rough plot
```{r}
aaa$row <- str_sub(aaa$Well, 1, 1)
aaa$column <- str_sub(aaa$Well, 2, 3)
#get a rough idea of what the samples are doing
gg <- ggplot(aaa[is.na(aaa$Mouse.ID) == F,], aes(x=Time, y=Abs570, color=Infected., label=Well))
gg + 
  geom_point()
#what you want to see is:
# 1. horizontal lines for all standards (pretty true)
# 2. linear increase in pos control w/ max abs at end of assay ~0.6 (this is what I normally get, and what examples plot in manual looks like)
# 3. samples should be higher than background if you're expecting arginase activity
# 4. if background absorbance is greater than 1.0, you should dilute the sample again
# 5. plots should be roughly linear. if not & if max abs is >1, consider diluting

#better plot with background absorption [[that is from the blank]] subtracted
gg <- ggplot(aaa, aes(x=Time, y=corrected, color=Infected.))

gg + 
  geom_point() +
  facet_wrap(~welltype)

#even better plot with background absorption [[that is from the sample]] subtracted
samples <- aaa[aaa$sampletype != 'standard',]
samples <- samples %>%
  group_by(Mouse.ID, Time, Day) %>%
  mutate(correctedcorrected = corrected[welltype == 'sample'] - corrected[welltype == 'background'])

#note that this does not remove the background points; do that here
samples <- samples[samples$welltype == 'sample',]

samples <- data.frame(samples)

gg <- ggplot(samples, aes(x=Time, y=correctedcorrected, color=Infected.))
gg + 
  geom_point()
```

##instructions say to plot corrected absorbance values for each standard as a function of the final concentration of arginase
```{r}
#if standards look pretty consistent over time, plot their averages
standards <- aaa[aaa$welltype == 'standard',]

standards_simple <- standards %>%
  group_by(Mouse.ID) %>%
  transmute(meanabs = mean(Abs570[Time==0], na.rm = TRUE),
            h2o2_conc = h2o2_conc)

standards_simple <- unique(standards_simple)

gg <- ggplot(standards_simple, aes(x=h2o2_conc, y=meanabs))
gg + geom_point(size=3)
#not bad!

#get equation from standard curve
x <- standards_simple$h2o2_conc
y <- standards_simple$meanabs - standards_simple$meanabs[1]

#get curve, extract components
curve <- lm(y~x)
summary(curve)
slope <- curve$coefficients['x']
intercept <- curve$coefficients['(Intercept)']

#plot to make sure they make sense
gg <- ggplot(standards_simple, aes(x=h2o2_conc, y=meanabs - meanabs[1]))
gg + 
  geom_point(size=3) +
  geom_abline(slope=slope,
              intercept=intercept)
#nice looks good

#now choose your timepoints. replot an earlier plot to choose a timerange over which the points increase in roughly linear fashion
gg <- ggplot(samples, aes(x=Time, y=correctedcorrected, color=Infected.))
gg + 
  geom_line(aes(group=Mouse.ID)) +
  geom_vline(xintercept = 8) +
  geom_vline(xintercept = 18) +
  facet_wrap(~Infected.)

########subset the data to just keep change in OD over that timerange ############
time1 <- 8
time2 <- 18

samples_overtime <- samples %>%
  group_by(Mouse.ID,Day,amt.sample.added) %>% #when using transmute, make sure you add all the columns you want to keep into 'group_by'
  transmute(deltaOD = correctedcorrected[Time == time2] - correctedcorrected[Time == time1],
         deltaTime = time2 - time1)

samples_overtime <- unique(samples_overtime)

####################################################################################

#calcuate h2o2 amount created for each of the samples, converting from deltaOD to umol of H2O2 produced using the standard curve values
#in other words, solve for x values for each sample using deltaOD values as y values
# x = (y - intercept) / slope
samples_overtime$h2o2_umol <- (samples_overtime$deltaOD - intercept) / slope

samples_overtime$h2o2permin <- samples_overtime$h2o2_umol / samples_overtime$deltaTime

#this is approximate for now, just to show that there is arginase activity
samples_overtime$ArgActivity <- samples_overtime$h2o2permin / samples_overtime$amt.sample.added

samples_overtime <- data.frame(samples_overtime)
```

#plot by treatment
```{r}
#add metadata
smd <- data.frame(read.xlsx('/Users/nicole/Desktop/Science/Schneider Lab/Malaria/_MouseStrains_x_Metabolites/strains_metabs_data_raw.xlsx'))
smd <- smd[c('Mouse.ID','SacDay','Strain','ALT','arginine','ornithine','citrulline')]
smd$Strain <- factor(smd$Strain, levels = resilience_levels)
samples_overtime2 <- merge(samples_overtime, smd, by.x = 'Mouse.ID', by.y = 'Mouse.ID', all.x = T)

#no B6 samples, so modify redbluePalette
redbluePalette_noB6 <- redbluePalette[c(1:4,6:8)]

gg <- ggplot(samples_overtime2[is.na(samples_overtime2$ALT) == F,], aes(x=ALT, y=ArgActivity, color=Strain))
arginase_ALT <- gg + 
  #geom_line(size=1) +
  geom_jitter() +
  scale_color_manual(values=redbluePalette_noB6) +
  guides(color=F) +
  labs(x = 'ALT (U/L)', y= 'Arginase Activity', color = 'Strain')
arginase_ALT

gg <- ggplot(samples_overtime2[is.na(samples_overtime2$ALT) == F,], aes(x=(ArgActivity), y=arginine, color=Strain))
arginase_arginine <- gg + 
  geom_smooth(color='black', span=1) +
  geom_jitter() +
  scale_color_manual(values=redbluePalette_noB6) +
  labs(x = 'Arginase Activity', y= 'Arginine', color = 'Strain')
arginase_arginine

setwd('/Users/nicole/Desktop/Science/Schneider Lab/Proposals, presentations, meetings/_Nicole/Papers/_Strains paper/figures/')

#ggsave(filename = 'prettyplots_4_strains_arginase_vs_arginine.pdf', arginase_arginine, width = 3, device = 'pdf', height = 1.75, units = 'in', useDingbats = F)
#ggsave(filename = 'prettyplots_4_strains_arginase_vs_ALT.pdf', arginase_ALT, width = 3, device = 'pdf', height = 1.75, units = 'in', useDingbats = F)

#write.xlsx(samples_overtime, '/Users/nicole/Desktop/Science/Schneider Lab/Malaria/Arginine/Argactivity_paperBatch4.xlsx')
```

Unit Definition:
1 Unit Arginase activity = amount of arginase that will generate 1.0 umol of H2O2 per minute at pH 8 at 37°C.
