---
title: "heat maps"
author: "Nicole Davis"
date: "February 16, 2020"
output: html_document
editor_options: 
  chunk_output_type: console
---

#prepare workspace
```{r, include=FALSE}
#load packages
library(ComplexHeatmap)
library(readxl)
library(psych)
library(reshape2)
library(ggplot2)
library(dplyr)
library(openxlsx)
library(ggrepel)
library(knitr)
library(stringr)
library(cowplot)
theme_set(theme_cowplot(font_size = 10))

redbluePalette <- c('red4','red3','red1','pink','steelblue1','royalblue2','mediumblue','navy')

resilience_levels <- c('PWK/PhJ','NOD/ShiLtJ','CAST/EiJ','A/J','C57BL/6','129S1/SvImJ','NZO/HILtJ','WSB/EiJ')

#set working directory
setwd('/Users/nicole/Desktop/Science/Schneider Lab/Malaria/R/correlations_Nicoles data_20200216/all data/')
```

#load longitudinal data (also retrieves metabolite names, which you need)
```{r, include=FALSE}
########load data here#########
strains_metabs_data <- data.frame(read_xlsx('strains_metabs_data_raw.xlsx'))
###########################################

strains_metabs_data$Parasite.Density <- strains_metabs_data$Parasite.Density/1000000
strains_metabs_data$Strain <- factor(strains_metabs_data$Strain, levels = resilience_levels)

strains_metabs_data$Infected_yn[strains_metabs_data$Infected_yn == TRUE] <- 'Inf'
strains_metabs_data$Infected_yn[strains_metabs_data$Infected_yn == FALSE] <- 'Uninf'

#get vectors that contain names of columns for each type of data
physcols <- c('RBC_Accuri','Blood.Glucose','Weight.Percent.Loss','Temperature.Delta','Parasitemia','Parasite.Density')
livercols <- c('AST','ALT')
cytocols <- c("EOTAXIN", "GSCF.CSF3", "GMCSF", "GROA", "IFNA", "IFNG", "IL10", "IL12P70", "IL13", "IL15.IL15R", "IL17A", "IL18", "IL1A", "IL1B", "IL2", "IL22", "IL23", "IL27", "IL28", "IL3", "IL31", "IL4", "IL5", "IL6", "IL9", "IP10", "LEPTIN", "LIF", "LIX", "MCSF", "MCP1", "MCP3", "MIP1A", "MIP1B", "MIP2", "RANTES", "TGFB", "TNFA", "VEGF")
flowcols <- c('Platelets', 'Early.Reticulocytes', 'Late.Reticulocytes', 'Erythrocytes', 'NK.Cells', 'B.Cells', 'Granulocytes', 'Granulocytes..Ly6C_pos', 'Granulocytes..Ly6C_neg', 'Monocytes', 'Monocytes..Ly6C_pos', 'Monocytes..Ly6C_neg', 'T.Cells', 'CD4.T.Cells', 'CD8.T.Cells', 'gd.T.Cells')

#first & last metabolite columns
start <- which(colnames(strains_metabs_data) == "X.16.or.17..methylstearate..a19.0.or.i19.0.")
end <- which(colnames(strains_metabs_data) == "xylose")

metabcols <- colnames(strains_metabs_data)[start:end]

median_cols <- c(physcols, flowcols, livercols, cytocols, metabcols)
```

#load extreme values
```{r, include=FALSE}
extremes <- data.frame(read.xlsx('/Users/nicole/Desktop/Science/Schneider Lab/Malaria/_MouseStrains_x_Metabolites/strains_metabs_extremes_alldays_normtob6.xlsx'))

extremes_inf <- extremes[extremes$Infected_yn == TRUE,]
extremes_inf$Infected_yn <- NULL
rownames(extremes_inf) <- extremes_inf$Strain
extremes_inf$Strain <- NULL #now you have just numeric columns
```

#load metabolites
```{r, include=FALSE}
####
metab_info <- read.xlsx('/Users/nicole/Desktop/Science/Schneider Lab/Malaria/R/correlations_Nicoles data_20200216/StrainsMetabs_metabolite_info.xlsx')

load('/Users/nicole/Desktop/Science/Schneider Lab/Malaria/R/correlations_Nicoles data_20200216/BIOCHEMICALS.match.Rdata')

metab_info <- left_join(metab_info, BIOCHEMICALS.match)

#remove duplicates
metab_infoG <- metab_info
metab_infoG <- metab_infoG[!(duplicated(metab_infoG$BIOCHEMICAL)),]
```

#remove bile acids
```{r, include=FALSE}
bileacids <- metab_infoG$BIOCHEMICAL.weirdformat[metab_infoG$SUB.PATHWAY %in% c('Primary Bile Acid Metabolism','Secondary Bile Acid Metabolism')]

metabs_keep <- metabcols[!(metabcols %in% bileacids)] #101
length(metabs_keep)

features_wanted <- c(physcols, livercols, metabs_keep)
```

#choose your data
#you want the data to include only columns containing numeric data
```{r, include=FALSE}
strains_metabs_inf <- strains_metabs_data[strains_metabs_data$Infected_yn == "Inf",]
strains_metabs_inf <- strains_metabs_inf[c('Mouse.ID', median_cols)]

#log transform the data
strains_metabs_inf[-1] <- log2(strains_metabs_inf[-1] - min(strains_metabs_inf[-1], na.rm=T) + 1)
```

#remove metabolites with zero standard deviation and unwanted physiology columns
#perform the correlation analysis
#adjust using Bonferroni
```{r, include=FALSE}
#####subset your columns here######
#df_to_plot is the dataframe you want to use
#features_wanted is a character vector of the 1st set of features you want to correlate with everything (these are the unknowns)
#correlation_features is a character vector of the features you want to correlate against everything (like health parameters)

get_corrs <- function(df_to_plot, features_wanted, correlation_features){
  inds <- which(colnames(df_to_plot) %in% features_wanted)  #prints your options
  zerosd <- sapply(df_to_plot[inds], FUN = sd, na.rm = TRUE)
  zerosd <- names(zerosd)[zerosd == 0]

  features_keep <- features_wanted[!(features_wanted %in% c(zerosd, 'RBC.Initial','Weight.Initial','Temperature.Initial','RBC'))] #RBC is a repeat of RBC_Accuri
  
  df <- df_to_plot[,colnames(df_to_plot) %in% features_keep]

  df <- df[rowSums(is.na(df)) != ncol(df), ] #removes rows that are entirely NAs
  
  df2 <- df[,colnames(df) %in% features_keep]
  
  health_params <- correlation_features
  
  #here is where you choose which things you want to correlate with what
  health <- df2[colnames(df2) %in% c(health_params)]
  params <- df2[!(colnames(df2) %in% c(health_params, 'AST','Parasitemia'))]
  
  corrs <- corr.test(health, params, use = 'pairwise', method = 'pearson', adjust = 'bonferroni', ci=FALSE)
  Rs <- data.frame(corrs[[1]])
  Rs$Var1 <- rownames(Rs)
  Rs <- melt(Rs, id.vars = 'Var1', variable.name='Var2', value.name = 'R')
  Rs$R2 <- Rs$R^2
  
  Ps <- data.frame(corrs[[4]])
  Ps$Var1 <- rownames(Ps)
  Ps <- melt(Ps, id.vars = 'Var1', variable.name='Var2', value.name = 'P')
  
  corrs <- left_join(Rs, Ps)
  
  #corrs <- corrs[corrs$Var1 != corrs$Var2,] #remove self-self correlations
  #corrs <- corrs[corrs$Var1 %in% health_params,] #select only correlations with health
  #corrs <- corrs[!(corrs$Var2 %in% health_params),] #remove health-health correlations
  
  corrs <- corrs %>%
    group_by(Var2) %>%
    mutate(Var2_bestR2 = max(R2, na.rm=TRUE),
           Var2_bestR = R[R2==Var2_bestR2][1])
  
  #corrs$P.adjust <- p.adjust(corrs$P, method='bonferroni') #even with FDR, the minimum P-value is 0.0735
  
  return(corrs)
}

health_params <- c('Parasite.Density','Parasitemia','RBC_Accuri','Temperature.Delta','Weight.Percent.Loss','Blood.Glucose','ALT', 'creatinine', 'urea', 'lactate')

corrs <- get_corrs(strains_metabs_inf, features_wanted, health_params)
```

#merge correlation data with metabolite metadata
```{r}
corrs_full <- merge(corrs, metab_info, by.x = 'Var2', by.y = 'BIOCHEMICAL.weirdformat', sort=FALSE, all.x = TRUE, all.y = TRUE)

#write.xlsx(corrs_full, '/Users/nicole/Desktop/Science/Schneider Lab/Malaria/R/correlations_Nicoles data_20200216/all data/healthcorrelations_nicolesdata.xlsx')

#David wants to know which metabolites correlate poorly with everything
corrs_best <- corrs_full
corrs_best <- corrs_best[corrs_best$R2 == corrs_best$Var2_bestR2,]

#write.xlsx(corrs_best, '/Users/nicole/Desktop/corrs for David.xlsx')
```

#plot the correlations
```{r, fig.height = 12, fig.width = 12}
#best flow correlation is with retics & FAD, hydroxystearate, and alpha-ketoglutarate; not super impressive on the volcano plot, but some are pretty good.

corrs_full$Number <- as.numeric(corrs_full$Var2)

corrs_full.volc <- corrs_full
corrs_full.volc <- corrs_full.volc[is.na(corrs_full.volc$Var1) == F,]
corrs_full.volc <- corrs_full.volc[corrs_full.volc$Var2 %in% metabcols,]

corrs_full.volc$Var1 <- str_replace(corrs_full.volc$Var1, 'RBC_Accuri', 'RBC')
corrs_full.volc$Var1 <- str_replace(corrs_full.volc$Var1, 'Temperature.Delta', 'Temperature')
corrs_full.volc$Var1 <- str_replace(corrs_full.volc$Var1, 'Weight.Percent.Loss', 'Weight Loss')
corrs_full.volc$Var1 <- str_replace(corrs_full.volc$Var1, 'Parasite.Density', 'Parasite Density')
corrs_full.volc$Var1[corrs_full.volc$Var1 %in% cytocols] <- 'Cytokine'
corrs_full.volc$Var1[corrs_full.volc$Var1 %in% flowcols] <- 'Cytometry'
corrs_full.volc$Var1[corrs_full.volc$Var1 == 'Blood.Glucose'] <- 'Blood Glucose'
corrs_full.volc$Var1[corrs_full.volc$Var1 == 'urea'] <- 'Urea'
corrs_full.volc$Var1[corrs_full.volc$Var1 == 'creatinine'] <- 'Creatinine'
corrs_full.volc$Var1[corrs_full.volc$Var1 == 'lactate'] <- 'Lactate'

corrs_full.volc <- corrs_full.volc[order(corrs_full.volc$R2, decreasing = T),]
corrs_full.volc <- corrs_full.volc %>%
  group_by(Var1) %>%
  mutate(rank = 1:length(Var1))

corrs_full.volc <- corrs_full.volc[order(corrs_full.volc$rank, decreasing = F),]

#just took best 50 correlations from each health metric & found their medians
metric.rankorder <- corrs_full.volc %>%
  group_by(Var1) %>%
  transmute(metric_median = median(R2, na.rm=T),
            metric_mean = mean(R2, na.rm=T),
            metric_top50median = median(R2[1:50], na.rm=T),
            metric_top50mean = mean(R2[1:50], na.rm=T)) %>%
  unique()
  
corrs_full.volc$Var1f <- factor(corrs_full.volc$Var1, levels = c('Urea','Temperature','ALT','Creatinine','Cytometry','RBC','Cytokine','Blood Glucose','Weight Loss', 'Parasitemia','Parasite Density','Lactate'))

#for shape
corrs_full.volc$rank2 <- 'one'
corrs_full.volc$rank2[corrs_full.volc$rank == 1] <- 'two'

#for pub
gg <- ggplot(corrs_full.volc[corrs_full.volc$Var1 %in% c('ALT','Parasitemia','RBC','Temperature','Weight Loss', 'Creatinine', 'Urea','Blood Glucose','Parasite Density'),], 
             aes(x=Var1f,
                 y=R2,
                 shape=rank2,
                 label=Number))

set.seed(1)
set.seed(2)

volcplot <- gg +
  geom_jitter(size = 2, width=0.3, height=0, fill='red') +
  geom_hline(linetype = 'dashed', color = 'blue', yintercept = 0.4) +
  scale_alpha_continuous(range = c(0.7, 1), guide='none') +
  scale_shape_manual(values = c(1,23)) +
  guides(shape=F) +
  #facet_wrap(~Var1f, nrow=1) +
  labs(x = '', y = bquote('R'^2), color = 'Health\nCorrelate') +
  theme(legend.position = 'bottom',
        axis.text.x = element_text(angle=45, hjust=1))

volcplot

maxwidth <- 6.875

#ggsave('prettyplots_3_volc.pdf', volcplot, height=3, width = maxwidth*2/3, units = 'in')
```

#plot individual correlations
```{r}
small <- strains_metabs_data[c('Mouse.ID','Strain')]
strains_metabs_inf <- merge(small, strains_metabs_inf, all.y = T, all.x = F)

theme_set(theme_cowplot(font_size=9))

gg <- ggplot(strains_metabs_inf, aes(x=urea, y=X3.hydroxypalmitoylcarnitine))
urea <- gg + geom_point(aes(color=Strain), size=1) + geom_smooth(method='lm',color='black') + 
  scale_x_continuous(breaks = c(4.4, 4.7, 5)) + 
  scale_y_continuous(breaks = c(4.5, 5, 5.5)) +
  scale_color_manual(values = redbluePalette) + labs(x='Urea',y='C16-OH') + theme(legend.position='none')
urea

gg <- ggplot(strains_metabs_inf, aes(x=ALT, y=arginine))
alt <- gg + geom_point(aes(color=Strain), size=1) + geom_smooth(method='lm',color='black', se=T) +
  scale_y_continuous(breaks = c(4.3, 4.4)) + scale_color_manual(values = redbluePalette) + labs(x='ALT',y='Arginine') + theme(legend.position='none')
alt

gg <- ggplot(strains_metabs_inf, aes(x=Temperature.Delta, y=pentadecanoylcarnitine..C15..))
temp <- gg + geom_point(aes(color=Strain), size=1) + geom_smooth(method='lm',color='black', se=T) +
  scale_y_continuous(breaks = c(4.5, 5)) +
  scale_x_continuous(breaks = c(3, 4)) +
  scale_color_manual(values = redbluePalette) + labs(x='Temperature',y='C15') + theme(legend.position='none')
temp

gg <- ggplot(strains_metabs_inf, aes(x=RBC_Accuri, y=N6.N6.dimethyllysine))
rbc <- gg + geom_point(aes(color=Strain), size=1) + geom_smooth(method='lm',color='black', se=T) + scale_color_manual(values = redbluePalette) +
  scale_x_continuous(breaks = c(4.4, 4.6, 4.8)) +
  scale_y_continuous(breaks = c(4.4, 4.6)) +
  labs(x='RBC',y='Dimethyllysine') + theme(legend.position='none')
rbc

gg <- ggplot(strains_metabs_inf, aes(x=creatinine, y=N.N.N.trimethyl.alanylproline.betaine..TMAP.))
creatinine <- gg + geom_point(aes(color=Strain), size=1) +
  geom_smooth(method='lm',color='black', se=T) + scale_color_manual(values = redbluePalette) + labs(x='Creatinine',y='TMAP') + theme(legend.position='none')
creatinine

gg <- ggplot(strains_metabs_inf, aes(x=creatinine, y=N6.carbamoylthreonyladenosine))
creatinine <- gg + geom_point(aes(color=Strain), size=1) +
  scale_x_continuous(breaks = c(4.4, 4.45)) +
  scale_y_continuous(breaks = c(4.4, 4.6, 4.8)) + geom_smooth(method='lm',color='black', se=T) + scale_color_manual(values = redbluePalette) + labs(x='Creatinine', y=bquote('t'^6~'a')) + theme(legend.position='none')
creatinine

gg <- ggplot(strains_metabs_inf, aes(x=Parasitemia, y=pipecolate))
parasitemia <- gg + geom_point(aes(color=Strain), size=1) +
  scale_x_continuous(breaks = c(4.33, 4.36)) +
  scale_y_continuous(breaks = c(4.4, 5, 5.6)) +
  geom_smooth(method='lm',color='black', se=T) + scale_color_manual(values = redbluePalette) + labs(x='Parasitemia',y='Pipecolate') + theme(legend.position='none')
parasitemia

gg <- ggplot(strains_metabs_inf, aes(x=Parasite.Density, y=N2.N2.dimethylguanosine))
parasite.density <- gg + geom_point(aes(color=Strain), size=1) +
  scale_y_continuous(limits = c(4.32,NA), breaks = c(4.4, 4.5)) +
  scale_x_continuous(breaks = c(4.35, 4.45)) +
  geom_smooth(method='lm',color='black', se=T) + scale_color_manual(values = redbluePalette) + labs(x='Parasite Density', y='Dimethylguanosine') + theme(legend.position='none')
parasite.density

gg <- ggplot(strains_metabs_inf, aes(x=Blood.Glucose, y=alpha.hydroxyisovalerate))
glucose <- gg + 
  geom_point(aes(color=Strain), size=1) +
  scale_y_continuous(breaks = c(4.4, 4.7, 5)) +
  geom_smooth(method='lm',color='black', se=T) + scale_color_manual(values = redbluePalette) + labs(x='Blood Glucose',y=bquote(alpha~'-hydroxyisovalerate')) + theme(legend.position='none')
glucose

gg <- ggplot(strains_metabs_inf, aes(x=Weight.Percent.Loss, y=X3.methylhistidine))
weight <- gg + geom_point(aes(color=Strain), size=1) +
  scale_x_continuous(breaks = c(4.31, 4.33)) +
  scale_y_continuous(breaks = c(4.4, 4.6)) +
  geom_smooth(method='lm',color='black', se=T) + scale_color_manual(values = redbluePalette) + labs(x='Weight Loss',y='3-methylhistidine') + theme(legend.position='none')
weight

#get legend
gg <- ggplot(metab1, aes(x=(MIP1B), y=log10(biliverdin + 10)))
cytokine2 <- gg + geom_point(aes(color=Strain), size=5) + geom_smooth(method='lm',color='black', se=T) + scale_color_manual(values = redbluePalette) + theme(legend.position='right') + labs(x='MIP1B', y='Biliverdin')
cytokine2

#plot with legend
leg <- get_legend(cytokine2 + theme(legend.justification = 'center', legend.direction = 'horizontal'))
#plot_grid(urea, temp, alt, creatinine, rbc, glucose, weight, parasitemia, cytokine, leg, nrow=2)

#save with appropriate text sizes
corr_plots <- plot_grid(plot_grid(urea, alt, temp, creatinine, rbc, parasitemia, glucose, parasite.density, weight, nrow = 3), leg, nrow=2, rel_heights = c(1, 0.15))

#ggsave('prettyplots_3_indiv_correlations.pdf', corr_plots, device = 'pdf', height=4.5, width=maxwidth*2/3, units='in', dpi=1200)
```