---
title: "sPLS-DA malaria"
author: "Nicole Davis"
date: "July 27, 2018"
output: html_document
editor_options: 
  chunk_output_type: console
---

#setup
```{r, include = FALSE}
bioc_packages <- c()
cran_packages <- c('mixOmics', 'ggplot2', 'reshape2', 'dplyr', 'openxlsx', 'readxl')

sapply(c(cran_packages, bioc_packages), require, character.only = TRUE)

theme_set(theme_bw())
options(stringsAsFactors = FALSE)
```

#load data
```{r, include = FALSE}
#strainsmetabs <- data.frame(read_xlsx('/Users/nicole/Desktop/Science/Schneider Lab/Malaria/_MouseStrains_x_Metabolites/strains_metabs_data_fc.xlsx'))

strainsmetabs <- data.frame(read_xlsx('/Users/nicole/Desktop/Science/Schneider Lab/Malaria/_MouseStrains_x_Metabolites/strains_metabs_data_raw.xlsx')) #these are scaled imputed from Metabolon

redbluePalette <- c('red4','red3','red1','pink','steelblue1','royalblue2','mediumblue','navy')

resilience_levels <- c('PWK/PhJ','A/J','NOD/ShiLtJ','CAST/EiJ','C57BL/6','129S1/SvImJ','NZO/HILtJ','WSB/EiJ') #disclaimer this is actually the order of max parasite density, not resilience, because resilience is harder to define until I've done more survival experiments

physcols <- c('RBC_Accuri','Blood.Glucose','Weight.Percent.Loss','Temperature.Delta','Parasitemia','Parasite.Density')
livercols <- c('AST','ALT')
cytocols <- c("EOTAXIN", "GSCF.CSF3", "GMCSF", "GROA", "IFNA", "IFNG", "IL10", "IL12P70", "IL13", "IL15.IL15R", "IL17A", "IL18", "IL1A", "IL1B", "IL2", "IL22", "IL23", "IL27", "IL28", "IL3", "IL31", "IL4", "IL5", "IL6", "IL9", "IP10", "LEPTIN", "LIF", "LIX", "MCSF", "MCP1", "MCP3", "MIP1A", "MIP1B", "MIP2", "RANTES", "TGFB", "TNFA", "VEGF")
cytocols_keep <- cytocols[!(cytocols %in% c('IL2','IL28'))]

flowcols <- c('Platelets', 'Early.Reticulocytes', 'Late.Reticulocytes', 'Erythrocytes', 'NK.Cells', 'B.Cells', 'Granulocytes', 'Granulocytes..Ly6C_pos', 'Granulocytes..Ly6C_neg', 'Monocytes', 'Monocytes..Ly6C_pos', 'Monocytes..Ly6C_neg', 'T.Cells', 'CD4.T.Cells', 'CD8.T.Cells', 'gd.T.Cells')

#load metabolite info
metab_info <- read.xlsx('/Users/nicole/Desktop/Science/Schneider Lab/Malaria/_MouseStrains_x_Metabolites/Metabolon/StrainsMetabs_metabolite_info.xlsx')

load('/Users/nicole/Desktop/Science/Schneider Lab/Malaria/_MouseStrains_x_Metabolites/BIOCHEMICALS.match.Rdata')

metab_info <- merge(metab_info, BIOCHEMICALS.match, by.x = 'BIOCHEMICAL', by.y = 'BIOCHEMICAL')

metabcols <- metab_info$BIOCHEMICAL.weirdformat

strainsmetabs_inf <- strainsmetabs[strainsmetabs$Infected_yn == TRUE,]

#do some subsetting
metabs_keep <- metab_info$BIOCHEMICAL.weirdformat 
metabs_rm <- metab_info$BIOCHEMICAL.weirdformat[metab_info$SUB.PATHWAY %in% c('Primary Bile Acid Metabolism','Secondary Bile Acid Metabolism')]
metabs_rm <- c(metabs_rm, 'hypoxanthine') #hypoxanthine is all NAs, so is removed. note, in earlier analysis it was not NAs, so not sure why it is now.

metabs_keep <- metabs_keep[!(metabs_keep %in% metabs_rm)]
params_keep <- c(metabs_keep)

strainsmetabs_inf <- strainsmetabs_inf[strainsmetabs_inf$SacDay %in% 7:9,]
  
gg <- ggplot(strainsmetabs[strainsmetabs$Strain %in% c('C57BL/6','CAST/EiJ'),], aes(x=SacDay, y=ALT, color=Strain))
gg + geom_point() + facet_wrap(~Infected_yn) 

#a simple choice:
# ALT low <= 2
# ALT hi > 3
```

#get the data you want for sPLS-DA
```{r, include=FALSE}
#sample subset
strainsmetabs_inf <- strainsmetabs_inf[is.na(strainsmetabs_inf$ALT) == FALSE,] #remove samples w/out ALT
strainsmetabs_inf <- strainsmetabs_inf[strainsmetabs_inf$Strain %in% c('C57BL/6','CAST/EiJ') & strainsmetabs_inf$SacDay %in% c(7:9),] #CAST and B6 only, at acute illness

######remove any parameters with sd == zero######
params_df <- strainsmetabs_inf[,colnames(strainsmetabs_inf) %in% params_keep]

zerosd <- sapply(params_df, FUN = sd, na.rm=TRUE)
zerosd <- names(zerosd)[zerosd == 0]
zerosd #just xanthine here

params_keep <- params_keep[!(params_keep %in% zerosd)]

strainsmetabs_inf <- strainsmetabs_inf[,colnames(strainsmetabs_inf) %in% c('Strain','Infected_yn','SacDay','ALT', params_keep)]
################################################

X = strainsmetabs_inf[4:ncol(strainsmetabs_inf)] #all numeric columns
X = X[colnames(X) != 'ALT'] #remove ALT column, that's the classification column
dim(X)
#log transform as in correlation analysis
X <- log2(X-min(X)+1)

#note: log10(1000) = 3, log10(100) = 2, log10(10) = 1
Y = log10(strainsmetabs_inf$ALT)
Y[Y > 3] <- 'high'
Y[Y <= 2] <- 'low'
Y[Y > 2 & Y <= 3] <- 'med'
Y

strainsmetabs_inf$ALT.class <- Y

Z = strainsmetabs_inf$SacDay
```

#summary of samples
```{r}
#liver damage inputs
liverdamage <- as.character(Y)
table(liverdamage, strainsmetabs_inf$Strain)
```

#do a pca
```{r}
ncomp <- 15

strains <- strainsmetabs_inf$Strain
alt <- strainsmetabs_inf$ALT.class
alt.f <- factor(strainsmetabs_inf$ALT.class, levels = c('low','med','high')) #this doesn't work at all with the plotting, the legend is way off

pca.malaria = pca(X, ncomp = ncomp, center = TRUE, scale = TRUE)

plot(pca.malaria, ncomp = ncomp)  # screeplot of the eingenvalues (explained variance per component)

plotIndiv(pca.malaria, group = alt, ellipse = TRUE, ind.names = FALSE, col=c('orange','blue','gray'),
          title = 'PCA', legend = T, legend.position = 'bottom')
```

Perhaps 4 components for PCA.

#perform PLS-DA

Class labels are resilience, ie whether a mouse strain is expected to die or live upon infection with malaria. A resilient animal survives.
```{r}
plsda.malaria <- plsda(X, Y, ncomp = 10)  # set ncomp to 10 for performance assessment later
```

```{r}
plotIndiv(plsda.malaria , comp = 1:2,
          group = alt, shape=resilience, ind.names = FALSE,
          ellipse = TRUE, legend = TRUE, title = 'PLSDA on malaria')
```

#Use `perf` to evaluate the performance of the fitted PLS-DA

Here I try M-fold cross-validation with 3 folds.

Note from function documentation that: "BER is appropriate in case of an unbalanced number of samples per class as it calculates the average proportion of wrongly classified samples in each class, weighted by the number of samples in each class. BER is less biased towards majority classes during the performance assessment.""
```{r}
# takes a couple of minutes to run
set.seed(2543) # for reproducibility, only when the `cpus' argument is not used
perf.plsda.malaria <- perf(plsda.malaria, validation = "loo",
                  progressBar = FALSE, auc = TRUE)

time <- Sys.time()

#save this
save(perf.plsda.malaria, file = paste0('/Users/nicole/Desktop/Science/2018_3summer/Stats 366 - Modern Stats for Modern Bio/Assignments/Project/perf.plsda.malaria_', time, '.Rdata'))
```

#
```{r}
load(paste0('/Users/nicole/Desktop/Science/2018_3summer/Stats 366 - Modern Stats for Modern Bio/Assignments/Project/perf.plsda.malaria_', time, '.Rdata'))

# perf.plsda.malaria$error.rate  # error rates
plot(perf.plsda.malaria, col = color.mixo(5:7), sd = TRUE, legend.position = "vertical")
```

```{r}
# run internally and saved

#set.seed(1234) # for reproducibility, only when the `cpus' argument is not used
# grid of possible keepX values that will be tested for each component
list.keepX <- c(1:10,  seq(20, 300, 10))

#ncomp selected based on results of the classification error rate plot
#in the tutorial, authors selected distance and measure (presumably) because it resulted in the lowest error. I selected maxdist based on low error rate. BER and overall error are about the same for high numbers of components.

t1 = proc.time()
tune.splsda.malaria <- tune.splsda(X, alt, ncomp = 3, validation = 'loo', progressBar = FALSE, test.keepX = list.keepX, cpus = 4)

t2 = proc.time()
running_time = t2 - t1; running_time # running time

error <- tune.splsda.malaria$error.rate # error rate per component for the keepX grid
error
ncomp2 <- tune.splsda.malaria$choice.ncomp$ncomp # optimal number of components based on t-tests
ncomp2 #if null, select 3 (based on PLS-DA earlier)
ncomp <- 3

time <- Sys.time()

save(tune.splsda.malaria, file =paste0('/Users/nicole/Desktop/Science/2018_3summer/Stats 366 - Modern Stats for Modern Bio/Assignments/Project/tune-malaria-sPLSDA_', time, '.RData'))
```

```{r}
#select number of features
select.keepX <- tune.splsda.malaria$choice.keepX[1:ncomp]
select.keepX

plot(tune.splsda.malaria, col = color.jet(3))

select.keepX[1:3] <- c(50,50,50)
```

#run the splsda
```{r}
splsda.malaria <- splsda(X, Y, ncomp = ncomp, keepX = select.keepX)

#for some reason, using the 'ggplot2' style with a factor for the group messes up the legend.

plotIndiv(splsda.malaria, comp = c(1,2),
          group = alt, ind.names = FALSE,
          ellipse = TRUE, legend = TRUE, legend.position = 'bottom', pch = c(2,2,2), col = c('blue','gray','orange'),
          title = 'CAST v. B6')
#save as 3.1 x 3.5 or so

plotIndiv(splsda.malaria, comp = c(1,3),
          group = alt, ind.names = FALSE,
          ellipse = TRUE, legend = TRUE,
          title = 'sPLS-DA on malaria, comp 1 & 3')

plotIndiv(splsda.malaria, comp = c(2,3),
          group = alt, ind.names = FALSE,
          ellipse = TRUE, legend = TRUE,
          title = 'sPLS-DA on malaria, comp 2 & 3')

auc.splsda = auroc(splsda.malaria, roc.comp = 2)
auc.splsda = auroc(splsda.malaria, roc.comp = 3)
```

#get legend

```{r}
set.seed(1)
set.seed(40) # for reproducibility, only when the `cpus' argument is not used
# takes about 1 min to run
perf.malaria <- perf(splsda.malaria, validation = "loo",
                   dist = 'all', progressBar = FALSE)

# perf.srbct  # lists the different outputs
perf.malaria$error.rate
plot(perf.malaria, col = color.mixo(c(5,2,1)))
```

A stability analysis is performed in which stable features are extracted. The analysis is performed using a lasso.

The authors of the package state in Le Cao et al. 2011 that "[40] proposed to keep in the selection only the variables that were selected in all subsamples, whereas [39] proposed to compute a relative selection frequency and keep the most stable variables in the selection. We chose to illustrate the latter option as we believe that the stability frequency, or probability, gives a better understanding of the number of stable discriminative variables that are selected in sPLS-DA."

The result of this is that: "The highly correlated variables will get a higher probability of being selected in each subsample, while the noisy variables will have a probability close to zero."

```{r}
par(mfrow=c(1,4))
plot(perf.malaria$features$stable[[1]], type = 'h', ylab = 'Stability',
     xlab = 'Features', main = 'Comp 1', las =2)
plot(perf.malaria$features$stable[[2]], type = 'h', ylab = 'Stability',
     xlab = 'Features', main = 'Comp 2', las =2)
plot(perf.malaria$features$stable[[3]], type = 'h', ylab = 'Stability',
     xlab = 'Features', main = 'Comp 3', las =2)
#plot(perf.malaria$features$stable[[3]], type = 'h', ylab = 'Stability',
#   xlab = 'Features', main = 'Comp 4', las =2)

par(mfrow=c(1,1))
```

Lots of stable features in component 1 and a few in the others. 

Le Cao et al. 2011: "We also noticed that once we reached too many dimensions (i.e. close K - 1), then the frequencies of all variables dropped, which clearly showed that sPLS-DA could not distinguish between discriminative variables and noisy variables any more (not shown)."

The axes on those plots are too small for inspection, let's look at a list of the stable ones.

```{r}
stable1 <- perf.malaria$features$stable[[1]]
names(stable1)[stable1 > 0.5]

stable2 <- perf.malaria$features$stable[[2]]
names(stable2)[stable2 > 0.5]

stable3 <- perf.malaria$features$stable[[3]]
names(stable3)[stable3 > 0.5]

stab1 <- data.frame(stable1,
                    comp = 1)
stab2 <- data.frame(stable2,
                    comp = 2)
stab3 <- data.frame(stable3,
                    comp = 3)

stab <- rbind(stab1, stab2, stab3)

stab2 <- merge(stab, metab_info, by.x = 'Var1', by.y = 'BIOCHEMICAL.weirdformat', all.x = T)

#write.xlsx(stab2, 'strains_liverdamage_castb6only_splsda_stablevariables2.xlsx')
```


```{r, fig.width=12, fig.height=12}
# here we match the selected variables to the stable features
ind.match = match(selectVar(splsda.malaria, comp = 1)$name,
                  names(perf.malaria$features$stable[[1]]))
#extract the frequency of selection of those selected variables
Freq = as.numeric(perf.malaria$features$stable[[1]][ind.match])

data.frame(selectVar(splsda.malaria, comp = 1)$value, Freq)

#get the loadings
tmp <- data.frame(splsda.malaria$loadings$X)

## ------------------------------------------------------------------------

spec_plotLoadings <- function(x){plotLoadings(splsda.malaria, comp = x, title = paste0('Loadings on comp ', x), contrib = 'max', method = 'median', size.legend = 0.5, ndisplay=30)}

sapply(1:ncomp, spec_plotLoadings)
```