---
title: "starvation metabolomics"
author: "Nicole Davis"
date: "3/10/2020"
output: html_document
editor_options: 
  chunk_output_type: console
---

This script analyzes data from Pietrocola et al. 2017, Autophagy, 
Metabolic effects of fasting on human and mouse blood in vivo

http://dx.doi.org/10.1080/15548627.2016.1271513

and from Andrianova et al. 2020, metabolites, Microbiome-Metabolome Signature of Acute Kidney Injury

https://doi.org/10.3390/metabo10040142

and from Ganeshan et al. 2019, Cell, Energetic Trade-Offs and Hypometabolic States Promote Disease Tolerance

https://doi.org/10.1016/j.cell.2019.01.050


```{r}
#load needed packages
library(dplyr)
library(stringr)
library(openxlsx)
library(magrittr)
library(reshape2)
library(ggplot2)
library(quantable)
library(scales)

#plotting themes
theme_set(theme_classic())
redbluePalette <- c('red1','royalblue2')

#load needed packages for heatmaps
library(ComplexHeatmap)
library(circlize)
```

#load data
```{r}
####### carnitines w/ AKI #######
carn <- data.frame(read.xlsx('/Users/nicole/Desktop/Science/Schneider Lab/Malaria/Metabolomics/carnitines/carnitines.xlsx'))

colnames(carn) <- c('metab','AKI','ncarns')
#################################

##### malarial mice ######
strainsMetabs <- data.frame(read.xlsx('/Users/nicole/Desktop/Science/Schneider Lab/Malaria/_MouseStrains_x_Metabolites/strains_metabs_data_fc.xlsx'))

strainsMetabs <- strainsMetabs[c(3,8,10,111:ncol(strainsMetabs))]

#the strains data is log2-transformed already, so undo that
strainsMetabs[4:ncol(strainsMetabs)] <- 2^strainsMetabs[4:ncol(strainsMetabs)]
##########################

####### starvation #######
starv <- data.frame(read.xlsx("/Users/nicole/Desktop/Science/Schneider Lab/Malaria/Metabolomics/starvation in mice and humans/starvation_carnitines.xlsx", sheet=1))

starv$FCStarv <- as.numeric(starv$FCStarv)
starv$FCStarv <- 2^starv$FCStarv #in this script, I undo all the log2 fold changes done in raw data file (if applicable) so I can redo it for all data together, later in script
##########################
```

#fold-change, strains
```{r}
#median calculations
strainsMetabs$SacDay[strainsMetabs$Infected_yn == FALSE] <- 0 ; strainsMetabs$Infected_yn <- NULL

#median value for B6 and CAST, just on d9
strainsMetabsMedian <- strainsMetabs %>%
  group_by(Strain, SacDay) %>%
  mutate_all(median, na.rm=TRUE) %>%
  unique()

rownames(strainsMetabsMedian) <- paste0('d', strainsMetabsMedian$SacDay, '_', strainsMetabsMedian$Strain)
strainsMetabsMedian <- data.frame(strainsMetabsMedian)

#just d9 for b6 and cast
strainsMetabsMedian <- strainsMetabsMedian[strainsMetabsMedian$Strain %in% c('C57BL/6','CAST/EiJ') & strainsMetabsMedian$SacDay == 9,]
strains.fc <- strainsMetabsMedian[3:ncol(strainsMetabsMedian)]
```

#merge with AKI data
```{r}
#now, you want to add two columns to the carnitines business - 1 for inf. B6 and 1 for inf. CAST
strains.fc <- data.frame(t(strains.fc))
colnames(strains.fc) <- c('BL6', 'CAST')

#merge
carn <- merge(carn, strains.fc, by.x = 'metab', by.y = 'row.names', all.x = TRUE, all.y = FALSE)
```

#merge starvation with carnitines df
```{r}
colnames(starv) <- c('metab', 'Starvation')

carn <- merge(carn, starv, by.x = 'metab', by.y = 'metab', all.x = T, all.y = F)
```

#add the Ganeshan LPS data
```{r}
#units are all micromolar
lps <- data.frame(read.xlsx('/Users/nicole/Desktop/Science/Schneider Lab/Malaria/Metabolomics/LPS in mice/Ganeshan et al. 2019 - table s1, acylcarnitines.xlsx', sheet = 'samplesmetabs'))
rownames(lps) <- lps$X1 ; lps$X1 <- NULL
lps <- data.frame(t(lps))

lps.meta <- data.frame(read.xlsx('/Users/nicole/Desktop/Science/Schneider Lab/Malaria/Metabolomics/LPS in mice/Ganeshan et al. 2019 - table s1, acylcarnitines.xlsx', sheet = 'samples'))
lps.meta$Sample.ID <- paste0('X',lps.meta$Name)

#get log2-foldchange
#merge
lps2 <- merge(lps.meta, lps, by.x = 'Sample.ID', by.y = 'row.names')
lps2 <- lps2[lps2$Temperature == 22,]

uninf.names <- lps2$Sample.ID[lps2$Treatment == 'Saline']

lps.uninf <- lps2[lps2$Sample.ID %in% uninf.names, 8:ncol(lps2)]
lps.uninf.values <- colMeans(lps.uninf)

lps.fc <- sweep(lps, 2, lps.uninf.values, '/')

inf.22.names <- lps2$Sample.ID[lps2$Treatment == '1 mg/kg LPS' & lps2$Temperature == 22]

lps.heat <- lps.fc[rownames(lps.fc) %in% inf.22.names,]
LPS <- data.frame(LPS = apply(lps.heat, 2, median))

carn <- merge(carn, LPS, by.x = 'metab', by.y = 'row.names', all.x = T)
rownames(carn) <- carn$Row.names; carn$Row.names <- NULL

carnM <- carn
rownames(carnM) <- carnM$metab ; carnM$metab <- NULL
carnM <- log2(carnM)

carnM <- carnM[order(carnM$ncarns, decreasing = F),]
carnM <- carnM[,c(4,3,6,1,5)] #removes ncarns, sorts
carnM <- data.frame(t(carnM))

#remove any column w/ 3 NAs or more, which roughly equates to 3 or fewer unique values
#get num NAs
my.fun <- function(x) {length(unique(x))}
numvals <- apply(carnM, 2, my.fun) #good, now subset to ones that have 4 or more unique values

carnM <- data.frame(t(carnM))
carnM$numvals <- numvals

carnM.sub <- carnM[carnM$numvals > 3,] ; carnM.sub$numvals <- NULL


#names are annoying, shorten them
rownames(carnM.sub)
newrownames <- c('C0','C2','C6-DC','C6','C8','C10:1','C10','C12','C12:1','C14:1','C14','C16:1','C16','C16-OH','C18:2','C18:1','C18','C18-OH','C20:4','C20','C20:1')

rownames(carnM.sub) <- newrownames
```

#plot as heatmap
```{r}
Heatmap(t(carnM.sub), col = colorRamp2(c(-3, 0, 3), c("blue", "white", "red")),
        show_heatmap_legend = T, 
        cluster_columns=F, 
        cluster_rows=F, 
        row_names_gp = gpar(fontsize = 20), 
        column_names_gp = gpar(fontsize = 20),
        heatmap_legend_param = list(title_gp = gpar(fontsize = 12), labels_gp =
                                      gpar(fontsize = 12)),
        name = 'log2\nfold-\nchange') #try colorRamp of -1 to 1
#save as 9x3.5

Heatmap(t(carnM.sub), col = colorRamp2(c(-3, 0, 3), c("blue", "white", "red")),
        show_heatmap_legend = T, 
        cluster_columns=F, 
        cluster_rows=F,
        row_names_gp = gpar(fontsize = 18), 
        column_names_gp = gpar(fontsize = 18), 
        name = 'log2\nfold-change') #try colorRamp of -1 to 1

col_fun = colorRamp2(c(0, 0.5, 1), c("blue", "white", "red"))
lgd = Legend(col_fun = col_fun, title = "foo")

Heatmap(t(carnM.sub), col = colorRamp2(c(-3, 0, 3), c("blue", "white", "red")),
        show_heatmap_legend = T, 
        cluster_columns=F, 
        cluster_rows=F,
        row_names_gp = gpar(fontsize = 18), 
        column_names_gp = gpar(fontsize = 18), 
        name = 'log2\nfold-change') #try colorRamp of -1 to 1
```
