---
title: "Malaria_prepdata"
author: "Nicole Davis"
date: "12/15/2017"
output: html_document
editor_options: 
  chunk_output_type: console
---

#use DOs/big screen data/follow-ups for parasitemia / physiology
#use DOs/follow-ups for survival

#prepare workspace
```{r}
library(readxl)
library(openxlsx)
library(ggplot2)
library(stringr)
library(dplyr)
library(reshape2)
library(cowplot)
library(survival)
library(survminer)
library(circlize)
theme_set(theme_bw())

#set colors & resilience levels
resilience_levels <- c('PWK/PhJ','A/J','NOD/ShiLtJ','CAST/EiJ','C57BL/6','129S1/SvImJ','NZO/HILtJ','WSB/EiJ') #set by average survival across experiments
resil <- c('C57BL/6','129S1/SvImJ','NZO/HILtJ','WSB/EiJ')
nonresil <- setdiff(resilience_levels, resil)

redbluePalette <- c('red4','red3','red1','pink','steelblue1','royalblue2','mediumblue','navy')
```

#"Batch" refers to a set of experiments, with one batch attempting to answer a specific question: the DOs are a batch by Avni and Victoria, the Strains x Metabs experiments are a batch by Nicole, and the next 3 experiments "follow ups" are another batch by Nicole to confirm survival phenotypes in the non-resilient strains

#"Experiment" refers to a cohort of mice infected on the same day. When assessing inter-experimental variation, use this variable

#load data, prep, plot
```{r}
surv.all <- data.frame(read.xlsx('/Dataset S1.xlsx', sheet = 'survival data'))

surv.all$Strain <- factor(surv.all$Strain, levels = resilience_levels)
surv.all$DeathDay[surv.all$DeathDay >= 15] <- 15

#variable is followed by the ~
fit <- survfit(Surv(DeathDay, Died.) ~ Strain, data = surv.all)
ggsurv <- ggsurvplot(fit, data = surv.all,
                     size = 1.25,
                     palette = redbluePalette,
                     xlab = "DPI", ylab = "\n Percent alive",
                     pval = F, # add p-value but not all p-values? can also do CI
                     legend = c("none"),
                     conf.int = F,
                     ggtheme = theme_cowplot(),
                     #font.x = 10,
                     #font.y = 10,
                     #font.tickslab = 9,
                     pval.size = 3,
                     censor = F)
survival <- ggsurv$plot
survival

#get a nice legend, the ggsurvplot ones are unpleasant
gg <- ggplot(surv.all, aes(x=DeathDay, y=Died., color=Strain))
leg <- gg + 
  geom_point() + 
  scale_color_manual(values = redbluePalette)

leg
```


#put together and save
```{r}
#get legend
legend <- get_legend(leg+ theme(legend.direction = "vertical", legend.justification="center", legend.box.just = "bottom")) # legend shenanigans

plot <- ggdraw(plot_grid(survival, legend, ncol = 2, rel_widths = c(1, 0.3)))
plot

#ggsave('prettyplots_tmp.pdf', plot, width=8, height=8, units="in", useDingbats=FALSE)

```
